---
title: "Final Project "
author: "Elie GNASSOU"
date: "5/4/2022"
output:
  word_document: default
  pdf_document: default
---

#**FINAL PROJECT DRAFT**

## Introduction 

**What**: For my final project, I originally wanted to find out the cost of health care in us prior covid and post covid to determine if COVID-19 has affected the price of healthcare at all or not and by how much compare to the previous years. But after looking for the data, I could not really find what I needed, so I have decided to put my on the focus the national spending to analyze hospital spending from 1960 to 2022. I have decided to analyze the spending on national insurance and find out if there is a correlation between them. 

 

**Why**: I think it would be interesting to see if there is an increase in the hospital spending and see if it affects the price of health insurance as all.

**How**: I will try to explore datasets about national health care and expenditure cost and use R studio to build a dashboard that show spending on hospital over the year and how much money spent on private insurance as well 

**Minimal scenario**: I explore the datasets but find just a little change in hospital spending and health insurance   

**Optimistic scenario**: I find the datasets and run my analysis and find that there is a substantial change in hospital and health insurance spending over the year   

**Ambitious scenario**: I explore the datasets and find that there is a relationship between hospital spending and health insurance. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##**TOPIC FROM CLASS**
Topic 1: Summarizing data 
 ## Using the steps from homework 1, 2,3 to examine data 
 ## Anaylyzing data by creating Histogram and plot
 
Topic 2: Github 
 
 
Topic 3: Linear Models  

 
 
 
 







```{r }
read.csv("nationalspending.csv")
```



```{r }
summary(nationalspending)
```


```{r }
dim(nationalspending)
```


```{r }
length(dim(nationalspending))
```


```{r }
names(nationalspending)
head(nationalspending)
tail(nationalspending)
```


```{r }
plot(nationalspending$Year, nationalspending$Hospital_spending)
```



```{r }
plot(nationalspending$Year, nationalspending$`Home_health_care ($)`)
```


```{r }
plot(nationalspending$Year, nationalspending$`Personal_healthcare ($)`)
```


```{r }
plot(nationalspending$Year, nationalspending$`Private_Health_Insurance ($)`)
```

```{r }
cor(nationalspending$Hospital_spending, nationalspending$Year)

```

```{r }
hist( x = nationalspending$Hospital_spending)
```



## CONCLUSION 
